---
title: "IODS final project"
author: "Sampsa Huttunen"
date: February 26, 2017
email: sampsa.huttunen@helsinki.fi
output:
  html_document:
    theme: lumen
    toc: true
    toc_depth: 2
    fig_caption: true
    fig_width: 6
    fig_height: 4
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

This final assignment analyzes the relation between extra-marital affairs (i.e. adultery or "cheating") with some other variables related to personal and marital life in the United States at the end of the 1970's. According to my results, the factor correlating most with the likelihood to have an extra-marital affair seems to be (maybe unsurprisingly) one's happiness to his/her marriage.  

## 1 Research Question

This assignment uses a dataset originally used in Ray C. Fair's 1977 (Fair 1977) study of extra-marital affairs to see what factors correlated with a tendency to cheat on one's spouse in the United States in the 1960s and 1970s.  

My hypothesis is that, while other factors may also play a role, not being happy in one's marriage is the strongest factor increasing the likelihood for an extra-marital relationship.

The statistical method applied in this assignment will be linear regression. It will be used to study the correlation between the number of an individual's extra-marital affairs ('nbaffairs', dependant variable) and other variables present in the data (independent variables).

## 2 Data

The dataset discussed here is a data of 601 answers of individuals to sex-related questions. It is a sample of a two surveys conducted in the United States: A survey by Psychology Today in 1969 and a survey by Redbook in 1974. Only employed people and people who were married and married for the first time were included in this dataset.

### 2.1 The Variables

##### The dataset contains the following variables:  
* sex - sex (male,female)  
* age - age  
* ym - number of years married
* child - children (no/yes) 
* religious - how religious from 1 (anti) to 5 (very)
* education - education:

  + 9.0 - Grade school  
  + 12.0 - High school
  + 14.0 - Some college
  + 16.0 - College graduate
  + 17.0 - Some graduate school
  + 18.0 - Master's degree
  + 20.0 - Ph.D., M.D., or other advanced degree.

* occupation - occupation, according to Hollingshead Index of Social Positions (ISP) Occupational Scale (with reversed numbering):

  + 7 - Higher executives of large concerns, proprietors, and major professionals  
  + 6 - Business managers, proprietors of medium-sized businesses, and lesser professionals  
  + 5 - Administrative personnel, owners of small businesses, and minor professionals
  + 4 - Clerical and sales workers, technicians, and owners of little businesses
  + 3 - Skilled manual employees
  + 2 - Machine operators and semiskilled employees
  + 1 - Unskilled employees.

* rate - self rating of marriage from 1 (very unhappy) to 5 (very happy)
* nbaffairs - number of affairs in the past year
  + 0 - none
  + 1 - Once  
  + 2 - Twice
  + 3 - 3 times
  + 7 - 4 to 10 times
  + 12 - Monthly or more

### 2.2 Data Wrangling

The original dataset 'Fair.csv' had an extra column with row index numbers. That column was omitted.  

The separator between values was also changed to ";" to make the CSV data comply and open neatly with the European Excel versions. The wrangled dataset was then written to a file 'affairs.csv' to be used in the analysis.  

### 2.3 The Files

The original dataset (Fair.csv) was downloaded from [https://vincentarelbundock.github.io/Rdatasets/csv/Ecdat/Fair.csv](https://vincentarelbundock.github.io/Rdatasets/csv/Ecdat/Fair.csv)  
The R script used to create the new dataset is at   [https://github.com/huttusam/IODS-final/blob/master/final-assignment.R](https://github.com/huttusam/IODS-final/blob/master/final-assignment.R)
and the 'affairs' dataset used in this assignment is at [https://github.com/huttusam/IODS-final/blob/master/data/affairs.csv](https://github.com/huttusam/IODS-final/blob/master/data/affairs.csv)

## 3 Analysis and Results

```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# Access libraries
library(dplyr)
library(ggplot2)
library(GGally)
library(stringr)
library(tidyverse)

# reading data from the wrangles CSV file to table 'affairs'
affairs <- read.table("data/affairs.csv", sep = ";", header = TRUE)

# printing a summary of affairs
summary(affairs)
# n= 601; 315 females, 286 males; mean age: 32.48 years
```

As the above summary indicates, 52% of the respondents in the data are female and 48% male. The mean age of the participants is 53,5 years and 71% of the participants have children. The average duration of marriage among the participants has been about 8 years.  

```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# printing a graphical summarization of affairs
gather(affairs) %>% ggplot(aes(value, alpha=0.3)) + facet_wrap("key", scales = "free") + geom_bar() + ggtitle("Barplots of all variables") +
  theme(plot.title = element_text(hjust = 0.5,size=16,face='bold'),legend.position="none")

```

When we look at the bar visualization of the data above, we can see that most participants have had at least some years of college education, are in administrative or managerial positions, are happy with their marriage, are somewhat religious, and have not had any extra-marital relationships. Who, then, has these relationships?  


```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
ggpairs(affairs, mapping = aes(col = sex, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20))) + ggtitle("All variables against each other") + theme(plot.title = element_text(hjust = 0.5,size=16,face='bold'))

```

The above summary shows that men and women don't differ very much in the number of extra-marital relations. The summary also indicates some interesting correlations with extra-marital affairs. There is a negative correlation between cheating and how people rate their marriage, some correlation between cheating and years in marriage, as well as a slight negative correlation between cheating and religiousness. Let's look at these pairs more closely:  


```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# Plotting nbaffairs vs rate
p1 <- ggplot(affairs, aes(x = rate, y = nbaffairs, col=sex)) +
  geom_point() + geom_smooth(method="lm") + 
  ggtitle("Happiness in Marriage vs Cheating")+
  theme(plot.title = element_text(hjust = 0.5,size=16,face='bold'),
        axis.title= element_text(hjust = 0.5,size=12))
p1

# fit a linear model
my_model1 <- lm(nbaffairs ~ rate, data = affairs)

# print out a summary of the model
summary(my_model1)
```

As we can see from the above linear regression plot, there is a clear negative correlation between the number of extra-marital affairs and the self-rated happiness of one's marriage. Although the data shows only a correlation between these two factors, the causality behind this correlation might run from experienced unhappiness to cheating. While this explanation feels plausible, further study is needed to confirm the direction of the causality.  

```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# Plotting nbaffairs vs ym
p2 <-ggplot(affairs, aes(x = ym, y = nbaffairs, col=sex)) +
  geom_point() + geom_smooth(method="lm") + 
  ggtitle("Number of Years Married vs Cheating")+
  theme(plot.title = element_text(hjust = 0.5,size=16,face='bold'),
        axis.title= element_text(hjust = 0.5,size=12))
p2

# fit a linear model
my_model2 <- lm(nbaffairs ~ ym, data = affairs)

# print out a summary of the model
summary(my_model2)

```

The number of years in marriage also seems to contribute to the likelihood of extra-marital relations. The bitter truth behind this might be that, over time, people get bored of their spouses and begin to seek some additional excitement to their (sex) life.  

```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# Plotting nbaffairs vs religious
p3 <-ggplot(affairs, aes(x = religious, y = nbaffairs, col=sex)) +
  geom_point() + geom_smooth(method="lm") + 
  ggtitle("Religiousness vs Cheating") +
  theme(plot.title = element_text(hjust = 0.5,size=16,face='bold'),
        axis.title= element_text(hjust = 0.5,size=12))
p3
# fit a linear model
my_model3 <- lm(nbaffairs ~ religious, data = affairs)

# print out a summary of the model
summary(my_model3)
``` 

Religiousness, on the other hand, seems to decrease the likelihood of extra-marital affairs. The type of the religion, the participant was affiliated with, was not in the data, but most of the world's religions condemn adultery. This, in turn, might decrease the religious person's willingness to practice extra-marital affairs.     


## References

Fair, R. (1977) "A note on the computation of the tobit estimator", Econometrica, 45, 1723-1727. [http://fairmodel.econ.yale.edu/rayfair/pdf/1978A200.PDF](http://fairmodel.econ.yale.edu/rayfair/pdf/1978A200.PDF)  


