---
title: "IODS final project"
author: "Sampsa Huttunen"
date: February 26, 2017
email: sampsa.huttunen@helsinki.fi
output:
  html_document:
    theme: lumen
    toc: true
    toc_depth: 2
    fig_caption: true
    fig_width: 6
    fig_height: 4
    code_folding: hide
---

## Abstract

An 'abstract' at the beginning of the page with a summary of your analysis  

## 1 Research Question

I am using a dataset originally used for Fair's 1977 study of extramarital affairs to see what factors correlated with a tendency to have extramarital affairs in the end of the 1960 in the United States. 

##### Script:  
* The R script used to create the new dataset (*with comments explaining the process*) is  [here](https://github.com/huttusam/IODS-final/blob/master/final-assignment.R)

## 2 Data and Variables

The dataset discussed here is a cross-section with 601 observations that are answers of individuals to sex-related questions. The dataset is a sample of a two surveys conducted in the United States: A survey by Psychology Today in 1969 and a survey by Redbook in 1974. Only employed people and people who were married and married for the first time were included in this dataset.

The original dataset (Fair.csv) was downloaded from [https://vincentarelbundock.github.io/Rdatasets/csv/Ecdat/Fair.csv](https://vincentarelbundock.github.io/Rdatasets/csv/Ecdat/Fair.csv)  

##### The dataset contains the following variables:  
* sex - sex (male,female)  
* age - age  
* ym - number of years married
* child - children (no/yes) 
* religious - how religious from 1 (anti) to 5 (very)
* education - education:
  + 9.0 - Grade school  
  + 12.0 - High school
  + 14.0 - Some college
  + 16.0 - College graduate
  + 17.0 - Some graduate school
  + 18.0 - Master's degree
  + 20.0 - Ph.D., M.D., or other advanced degree.
* occupation - occupation, according to Hollingshead Index of Social Positions (ISP) Occupational Scale (with reversed numbering):
  + 7 - Higher executives of large concerns, proprietors, and major professionals  
  + 6 - Business managers, proprietors of medium-sized businesses, and lesser professionals  
  + 5 - Administrative personnel, owners of small businesses, and minor professionals
  + 4 - Clerical and sales workers, technicians, and owners of little businesses
  + 3 - Skilled manual employees
  + 2 - Machine operators and semiskilled employees
  + 1 - Unskilled employees.
* rate - self rating of marriage from 1 (very unhappy) to 5 (very happy)
* nbaffairs - number of affairs in the past year.

The dataset was originally assembled and used in his research by Fair (1977).  

##### Data Wrangling

The original dataset 'Fair.csv' had an extra column with row index numbers. That column was omitted. The sparator between values was also changed to ";" to make the CSV data comply and open neatly with the European Excel versions. The wrangled dataset was then written to a file 'affairs.csv' to be used in the analysis.  

## 4 Analysis

Visually clear and interesting explorations of the variables of interest in the data, from the point of view of your research question. Include interpretations of the distributions and relationships of the variables. Use captions to draw the reader's focus on the interesting parts of your tables and graphics.  

First, let's look at the a summary of the data:

```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# Access libraries
library(dplyr)
library(ggplot2)
library(GGally)
library(stringr)
library(tidyverse)

# reading data from the wrangles CSV file to table 'affairs'
affairs <- read.table("data/affairs.csv", sep = ";", header = TRUE)

# printing a summary of affairs
summary(affairs)
# n= 601; 315 females, 286 males; mean age: 32.48 years
```

As we can see from the summary, about 52% of the respondents were femail and 48% male. The mean age of the participants was 53,5 years and about 71% of them had children. The average duration of marriage among the participants had been about 8 years.  


```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# printing a graphical summarization of affairs
gather(affairs) %>% ggplot(aes(value,fill="#99999",alpha=0.3)) + facet_wrap("key", scales = "free") + geom_bar() + ggtitle("Bars of affairs data") +
  theme(plot.title = element_text(hjust = 0.5,size=16,face='bold'),legend.position="none")

ggpairs(affairs, mapping = aes(col = sex, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20))) +
  ggtitle("affairs-data against each other") +
  theme(plot.title = element_text(hjust = 0.5,size=20,face='bold'))
```

When we look at a bar visualization of the data, we can see that most participants have had at least some years of college education, have had no extra-marital relationships, are in managerial or administrative positions, are happy with their marriage, and are somewhat religious.  

```{r echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
# Creating logical column 'high_cheat': TRUE if number of extra-marital relations is more than 1
affairs_ <- mutate(affairs, high_cheat = nbaffairs > 1)

```

## 5 Methodology

Brief description of the method you are using in your own words.  

## 6 Results

Presentation of the results of your analysis including visualizations and summaries and a thorough interpretation of the results including a validation analysis of the method.  

## 7 Conclusions and Discussion


## References
Fair, R. (1977) "A note on the computation of the tobit estimator", Econometrica, 45, 1723-1727. [http://fairmodel.econ.yale.edu/rayfair/pdf/1978A200.PDF](http://fairmodel.econ.yale.edu/rayfair/pdf/1978A200.PDF)  




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
